Forcasting US Elections using PollsofPolls

### 2024-10-21 paper/paper.qmd
Made some changes.

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(palmerpenguins)
```

@@

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

@@

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```
We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

@@

# Pollster

# Idealized Methodology

>>>>

---
title: Forcasting US Elections using PollsofPolls
subtitle: "My subtitle if needed"
author: 
  - Shamayla Durrin
  - Denise Chang 
  - Krishna Kumar
thanks: "Code and data are available at: [https://github.com/krishnak30/US_elections](https://github.com/krishnak30/US_elections)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(arrow)
library(here)
library(ggplot2)
analysis_data <- read_parquet(here("data/02-analysis_data/analysis_data.parquet"))
```

@@

# Polls of Polls to Forcast Winning Candidate

## Forecasting Approach

The polls of polls methodology is widely used in election prediction as it aggregates multiple polls to provide a more reliable estimate of voter support, rather than relying on any single poll. The goal is to reduce errors and biases present in individual polls by using a weighted average of many different polls. In our approach, we calculate the weights based on several important variables. Sample size is included because larger polls are generally more reliable and have smaller margins of error. Pollscore (or pollster rating) reflects the empirical quality of a pollster, balancing both accuracy and bias, allowing us to give more weight to those with a proven track record. We also account for the pollscope, differentiating between national and state-level polls, as state polls may offer more localized insights into voter sentiment. We also use recency as more recent polls give us a more accurate estimate of voter sentiment.

However, we acknowledge that even with these adjustments, there may still be underlying biases from individual pollsters toward the candidates. To account for this, we will select a baseline pollster with a strong rating, high transparency, and minimal historical error, and regress the support percentage (pct) on pollster, including the same variables as controls whcih was used in the weighting method, such as sample size, pollscore and pollscope.Additionally, we will control for the state of the poll, as support levels can vary regionally. Therefore each pollster's coefficient in the regression model will represent the bias of that pollster relative to the baseline pollster. Controlling for these factors is crucial because it ensures that the differences between pollsters are not driven by other factors, such as differences in sample size or the types of polls they conduct, but by their inherent bias in support estimates. Once we have calculated the weights, we will use them to generate a weighted average of support for Kamala Harris and Donald Trump, ultimately comparing their weighted support to forecast the likely winner of the election.

## **Exploration of Control Variables**

### Investigating the Effect of Pollscore on Candidate Support

```{r}

# Define pollscore ranges (you can adjust the binwidth if needed)
cleaned_data <- cleaned_data %>%
  mutate(pollscore_range = cut(pollscore, breaks = seq(-1.5, 1, by = 0.3), include.lowest = TRUE))

# Calculate the average support for each candidate within each pollscore range
average_support <- cleaned_data %>%
  group_by(pollscore_range, candidate_name) %>%
  summarise(avg_support = mean(pct, na.rm = TRUE)) %>%
  ungroup()

# Create a bar plot
ggplot(average_support, aes(x = pollscore_range, y = avg_support, fill = candidate_name)) +
  geom_bar(stat = "identity", position = "dodge") +  # Dodge to separate bars for each candidate
  scale_fill_manual(values = c("Donald Trump" = "indianred", "Kamala Harris" = "steelblue")) +  # Custom colors
  labs(title = "Average Support for Candidates by Pollscore Range",
       x = "Pollscore Range",
       y = "Average Support (%)",
       fill = "Candidate") +
  theme_minimal() +  # Minimal theme
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5)
  )

```

### Investigating the Effect of Poll Recency on Candidate Support

```{r}
#| warning: false
#| message: false
#| echo: false
# Custom colors for candidates
trump_color <- "indianred"
harris_color <- "steelblue"

# Create the plot with the requested adjustments (recency vs support)
ggplot(analysis_data, aes(x = recency, y = pct, color = candidate_name)) +
  geom_point(size = 0.5, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) + # Smaller dots with slight transparency
  scale_color_manual(values = c("Donald Trump" = trump_color, "Kamala Harris" = harris_color)) +  # Custom colors
  theme_minimal() +  # Remove grey background
  theme(
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank(),  # Remove minor gridlines
    legend.title = element_blank(),  # Remove "candidate_name" as title
    legend.position = "bottom",  # Move legend to the bottom
    plot.title = element_text(hjust = 0.5),  # Center the title
    axis.line = element_line(color = "black"),  # Add axis lines
    panel.border = element_blank(),  # Remove panel border
    axis.line.y.right = element_line(color = "black"),  # Add right spine for Kamala Harris
    axis.line.x.top = element_blank()  # Remove top spine
  ) +
  labs(title = "Recency vs Support for Kamala Harris and Donald Trump",
       x = "Poll Recency (days since poll end date)", 
       y = "Support (%)") +
  facet_wrap(~ candidate_name, scales = "free_y")  # Separate plots for each candidate

```

### Investigate the effect of sample size on support estimates

```{r}
#| warning: false
#| message: false
#| echo: false

# Custom colors for candidates
trump_color <- "indianred"
harris_color <- "steelblue"

# Create the plot with the requested adjustments
ggplot(analysis_data, aes(x = sample_size, y = pct, color = candidate_name)) +
  geom_point(size = 0.5, alpha = 0.9) + # Smaller dots with slight transparency
  geom_smooth(method = "lm", se = FALSE)+

  scale_color_manual(values = c("Donald Trump" = trump_color, "Kamala Harris" = harris_color)) +  # Custom colors
  theme_minimal() +  # Remove grey background
  theme(
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank(),  # Remove minor gridlines
    legend.title = element_blank(),  # Remove "candidate_name" as title
    legend.position = "bottom",  # Move legend to the bottom
    plot.title = element_text(hjust = 0.5),  # Center the title
    axis.line = element_line(color = "black"),  # Add axis lines
    panel.border = element_blank(),  # Remove panel border
    axis.line.y.right = element_line(color = "black"),  # Add right spine for Kamala Harris
    axis.line.x.top = element_blank()  # Remove top spine
  ) +
  labs(title = "Sample Size vs Support for Kamala Harris and Donald Trump",
       x = "Sample Size", 
       y = "Support (%)") +
  facet_wrap(~ candidate_name, scales = "free_y")
```

### Distribution of support for Candiates by State

```{r}
# Calculate mean support for each candidate in each state
state_support <- analysis_data %>%
  group_by(state, candidate_name) %>%
  summarise(mean_support = mean(pct)) %>%
  ungroup()

# Pivot the data to have one row per state with separate columns for Harris and Trump
state_support_wide <- state_support %>%
  pivot_wider(names_from = candidate_name, values_from = mean_support)
```

```{r}
# Calculate mean support for each candidate in each state
state_support <- analysis_data %>%
  group_by(state, candidate_name) %>%
  summarise(mean_support = mean(pct))

# Calculate the proportion of support for Kamala Harris
state_support <- state_support %>%
  mutate(harris_proportion = Kamala_Harris / (Kamala_Harris + Donald_Trump))

```

@@

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

@@

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```
